A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)

### 2019-11-13 Simulation/Writeup/preprint.tex
Draft research report for peer feedback research seminar

From the simulation diagnostics (...) it appears that as little as three iterations is sufficient to draw valid inference. Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate 20-30 iterations. Apparently, they are not perfect for MI data.

Moreover, $\widehat{R}$ could  theoretically not be smaller than one, yet it happened several times in this study (see online appendix XYZ). How could $\widehat{R}$ smaller than 1 occur? The number of simulations is smaller than in 'regular' MCMC processes. Therefore, the '$(n-1/n)$' correction factor can influence the estimated potential scale reduction factor. This downwards bias is in the opposite direction than expected.

% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Tue Nov 12 17:43:11 2019
\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrr}
  \hline
It. & Bias & Emp. SE & CI width & Cov. rate & $\widehat{R}$ & Auto-corr. \\ 
  \hline
  1 & -0.135 & 0.115 & 0.637 & 0.930 &  &   \\ 
  2 & -0.088 & 0.125 & 0.691 & 0.990 & 1.499 & -0.500 \\ 
  3 & -0.090 & 0.112 & 0.623 & 0.950 & 1.200 & -0.658 \\ 
  4 & -0.093 & 0.116 & 0.645 & 0.980 & 1.146 & -0.738 \\ 
  5 & -0.087 & 0.108 & 0.599 & 0.980 & 1.098 & -0.711 \\ 
  6 & -0.092 & 0.116 & 0.644 & 0.950 & 1.090 & -0.674 \\ 
  7 & -0.095 & 0.122 & 0.675 & 1.000 & 1.063 & -0.588 \\ 
  8 & -0.099 & 0.113 & 0.626 & 0.970 & 1.058 & -0.523 \\ 
  9 & -0.095 & 0.115 & 0.639 & 1.000 & 1.044 & -0.519 \\ 
  10 & -0.090 & 0.111 & 0.618 & 0.990 & 1.048 & -0.414 \\ 
  15 & -0.093 & 0.109 & 0.604 & 0.950 & 1.020 & -0.339 \\ 
  20 & -0.098 & 0.109 & 0.606 & 0.980 & 1.023 & -0.234 \\ 
  25 & -0.100 & 0.113 & 0.626 & 0.990 & 1.016 & -0.091 \\ 
  30 & -0.091 & 0.118 & 0.654 & 0.960 & 1.014 & -0.153 \\ 
  40 & -0.088 & 0.117 & 0.650 & 0.990 & 1.009 & -0.070 \\ 
  50 & -0.087 & 0.119 & 0.662 & 0.990 & 1.009 & -0.029 \\ 
  % 11 & -0.100 & 0.127 & 0.704 & 0.970 & 1.046 & -0.360 \\ 
  % 12 & -0.096 & 0.115 & 0.641 & 0.980 & 1.032 & -0.354 \\ 
  % 13 & -0.098 & 0.120 & 0.664 & 0.960 & 1.034 & -0.328 \\ 
  % 14 & -0.094 & 0.113 & 0.630 & 0.990 & 1.030 & -0.353 \\ 
  % 15 & -0.093 & 0.109 & 0.604 & 0.950 & 1.020 & -0.339 \\ 
  % 16 & -0.094 & 0.114 & 0.630 & 0.980 & 1.025 & -0.272 \\ 
  % 17 & -0.091 & 0.116 & 0.645 & 0.980 & 1.028 & -0.243 \\ 
  % 18 & -0.091 & 0.116 & 0.645 & 0.990 & 1.023 & -0.179 \\ 
  % 19 & -0.087 & 0.114 & 0.630 & 0.980 & 1.024 & -0.281 \\ 
  % 20 & -0.098 & 0.109 & 0.606 & 0.980 & 1.023 & -0.234 \\ 
  % 21 & -0.102 & 0.116 & 0.646 & 0.960 & 1.017 & -0.192 \\ 
  % 22 & -0.095 & 0.114 & 0.633 & 0.980 & 1.016 & -0.119 \\ 
  % 23 & -0.088 & 0.107 & 0.595 & 0.980 & 1.017 & -0.246 \\ 
  % 24 & -0.097 & 0.112 & 0.621 & 0.980 & 1.016 & -0.148 \\ 
  % 25 & -0.100 & 0.113 & 0.626 & 0.990 & 1.016 & -0.091 \\ 
  % 26 & -0.096 & 0.113 & 0.628 & 0.980 & 1.017 & -0.177 \\ 
  % 27 & -0.096 & 0.117 & 0.650 & 0.980 & 1.014 & -0.167 \\ 
  % 28 & -0.102 & 0.116 & 0.645 & 0.960 & 1.015 & -0.038 \\ 
  % 29 & -0.098 & 0.114 & 0.632 & 0.990 & 1.016 & -0.137 \\ 
  % 30 & -0.091 & 0.118 & 0.654 & 0.960 & 1.014 & -0.153 \\ 
  % 31 & -0.098 & 0.115 & 0.639 & 0.980 & 1.012 & -0.101 \\ 
  % 32 & -0.096 & 0.113 & 0.626 & 0.940 & 1.011 & -0.095 \\ 
  % 33 & -0.089 & 0.117 & 0.652 & 0.970 & 1.011 & -0.088 \\ 
  % 34 & -0.093 & 0.115 & 0.639 & 0.960 & 1.013 & -0.061 \\ 
  % 35 & -0.103 & 0.115 & 0.638 & 0.970 & 1.011 & -0.126 \\ 
  % 36 & -0.090 & 0.113 & 0.630 & 0.980 & 1.010 & -0.161 \\ 
  % 37 & -0.098 & 0.122 & 0.675 & 0.990 & 1.011 & -0.070 \\ 
  % 38 & -0.095 & 0.118 & 0.656 & 0.970 & 1.011 & -0.119 \\ 
  % 39 & -0.091 & 0.121 & 0.670 & 0.980 & 1.009 & -0.100 \\ 
  % 40 & -0.088 & 0.117 & 0.650 & 0.990 & 1.009 & -0.070 \\ 
  % 41 & -0.095 & 0.117 & 0.648 & 0.960 & 1.010 & -0.148 \\ 
  % 42 & -0.095 & 0.118 & 0.657 & 0.990 & 1.009 & -0.117 \\ 
  % 43 & -0.095 & 0.116 & 0.645 & 0.980 & 1.009 & -0.055 \\ 
  % 44 & -0.101 & 0.125 & 0.692 & 0.980 & 1.009 & -0.068 \\ 
  % 45 & -0.085 & 0.117 & 0.651 & 0.980 & 1.008 & -0.086 \\ 
  % 46 & -0.092 & 0.117 & 0.649 & 0.970 & 1.008 & -0.048 \\ 
  % 47 & -0.095 & 0.115 & 0.638 & 0.980 & 1.007 & -0.069 \\ 
  % 48 & -0.096 & 0.112 & 0.620 & 0.970 & 1.009 & -0.076 \\ 
  % 49 & -0.090 & 0.108 & 0.601 & 0.980 & 1.009 & -0.083 \\ 
  % 50 & -0.087 & 0.119 & 0.662 & 0.990 & 1.009 & -0.029 \\ 
   \hline
\end{tabular}
\label{subsetresults}
\end{table}

@@

"The mixture-of-sequences variance, V, should stabilize as a function of n. (Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)" \cite[p~438]{broo98}.

>>>>

Evaluate simulation diagnostics one by one: bias, empirical SE, confidence interval width, and coverage rate. Then evaluate convergence diagnostics $\widehat{R}$ and auto-correlation.

%% --------------------------------
\subsection{Simulation diagnostics}
Bias. After 1 or 2 iterations bias fluctuates within a narrow range of the average bias across all simulation conditions.

Empirical standard error (SE).

Confidence interval width. Does not show a trend across the simulation conditions. [Is related to empirical SE??]

Coverage rate. Is more or less stable after one iteration. But higher percentages then the expected 95\% \cite{neym43}.

%% --------------------------------
\subsection{Convergence diagnostics}

$\widehat{R}$. Steep drop down up-to iteration 30-40, then stable. Conventional threshold $\widehat{R} < 1.2$ reached after three iterations, $\widehat{R} < 1.1$ after four iterations, $\widehat{R} < 1.01$ after 38 iterations (see online appendix).

Auto-correlation. Auto-correlation is dangerous when positive. These auto-correlations are negative. Still, we want the iterations to be stable. Low auto-correlations

% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Tue Nov 12 17:43:11 2019
\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrrr}
  \hline
  &\multicolumn{4}{c}{Simulation Diagnostics}& & \multicolumn{2}{c}{Convergence Diagn.} \\
  \cline{2-5} \cline{7-8}
It. & Bias & Emp. SE & CI width & Cov. rate & & $\widehat{R}$ & Auto-corr. \\ 
  \hline
  1 & -0.135 & 0.115 & 0.637 & 0.930 &  & & \\ 
  2 & -0.088 & 0.125 & 0.691 & 0.990 & & 1.499 & -0.500 \\ 
  3 & -0.090 & 0.112 & 0.623 & 0.950 & &  1.200 & -0.658 \\ 
  4 & -0.093 & 0.116 & 0.645 & 0.980 &  & 1.146 & -0.738 \\ 
  5 & -0.087 & 0.108 & 0.599 & 0.980 &  & 1.098 & -0.711 \\ 
  6 & -0.092 & 0.116 & 0.644 & 0.950 & &  1.090 & -0.674 \\ 
  7 & -0.095 & 0.122 & 0.675 & 1.000 & &  1.063 & -0.588 \\ 
  8 & -0.099 & 0.113 & 0.626 & 0.970 & &  1.058 & -0.523 \\ 
  9 & -0.095 & 0.115 & 0.639 & 1.000 & &  1.044 & -0.519 \\ 
  10 & -0.090 & 0.111 & 0.618 & 0.990 & &  1.048 & -0.414 \\ 
  15 & -0.093 & 0.109 & 0.604 & 0.950 & &  1.020 & -0.339 \\ 
  20 & -0.098 & 0.109 & 0.606 & 0.980 & &  1.023 & -0.234 \\ 
  25 & -0.100 & 0.113 & 0.626 & 0.990 &  & 1.016 & -0.091 \\ 
  30 & -0.091 & 0.118 & 0.654 & 0.960 & &  1.014 & -0.153 \\ 
  40 & -0.088 & 0.117 & 0.650 & 0.990 &  & 1.009 & -0.070 \\ 
  50 & -0.087 & 0.119 & 0.662 & 0.990 &  & 1.009 & -0.029 \\ 
  % 11 & -0.100 & 0.127 & 0.704 & 0.970 & 1.046 & -0.360 \\ 
  % 12 & -0.096 & 0.115 & 0.641 & 0.980 & 1.032 & -0.354 \\ 
  % 13 & -0.098 & 0.120 & 0.664 & 0.960 & 1.034 & -0.328 \\ 
  % 14 & -0.094 & 0.113 & 0.630 & 0.990 & 1.030 & -0.353 \\ 
  % 15 & -0.093 & 0.109 & 0.604 & 0.950 & 1.020 & -0.339 \\ 
  % 16 & -0.094 & 0.114 & 0.630 & 0.980 & 1.025 & -0.272 \\ 
  % 17 & -0.091 & 0.116 & 0.645 & 0.980 & 1.028 & -0.243 \\ 
  % 18 & -0.091 & 0.116 & 0.645 & 0.990 & 1.023 & -0.179 \\ 
  % 19 & -0.087 & 0.114 & 0.630 & 0.980 & 1.024 & -0.281 \\ 
  % 20 & -0.098 & 0.109 & 0.606 & 0.980 & 1.023 & -0.234 \\ 
  % 21 & -0.102 & 0.116 & 0.646 & 0.960 & 1.017 & -0.192 \\ 
  % 22 & -0.095 & 0.114 & 0.633 & 0.980 & 1.016 & -0.119 \\ 
  % 23 & -0.088 & 0.107 & 0.595 & 0.980 & 1.017 & -0.246 \\ 
  % 24 & -0.097 & 0.112 & 0.621 & 0.980 & 1.016 & -0.148 \\ 
  % 25 & -0.100 & 0.113 & 0.626 & 0.990 & 1.016 & -0.091 \\ 
  % 26 & -0.096 & 0.113 & 0.628 & 0.980 & 1.017 & -0.177 \\ 
  % 27 & -0.096 & 0.117 & 0.650 & 0.980 & 1.014 & -0.167 \\ 
  % 28 & -0.102 & 0.116 & 0.645 & 0.960 & 1.015 & -0.038 \\ 
  % 29 & -0.098 & 0.114 & 0.632 & 0.990 & 1.016 & -0.137 \\ 
  % 30 & -0.091 & 0.118 & 0.654 & 0.960 & 1.014 & -0.153 \\ 
  % 31 & -0.098 & 0.115 & 0.639 & 0.980 & 1.012 & -0.101 \\ 
  % 32 & -0.096 & 0.113 & 0.626 & 0.940 & 1.011 & -0.095 \\ 
  % 33 & -0.089 & 0.117 & 0.652 & 0.970 & 1.011 & -0.088 \\ 
  % 34 & -0.093 & 0.115 & 0.639 & 0.960 & 1.013 & -0.061 \\ 
  % 35 & -0.103 & 0.115 & 0.638 & 0.970 & 1.011 & -0.126 \\ 
  % 36 & -0.090 & 0.113 & 0.630 & 0.980 & 1.010 & -0.161 \\ 
  % 37 & -0.098 & 0.122 & 0.675 & 0.990 & 1.011 & -0.070 \\ 
  % 38 & -0.095 & 0.118 & 0.656 & 0.970 & 1.011 & -0.119 \\ 
  % 39 & -0.091 & 0.121 & 0.670 & 0.980 & 1.009 & -0.100 \\ 
  % 40 & -0.088 & 0.117 & 0.650 & 0.990 & 1.009 & -0.070 \\ 
  % 41 & -0.095 & 0.117 & 0.648 & 0.960 & 1.010 & -0.148 \\ 
  % 42 & -0.095 & 0.118 & 0.657 & 0.990 & 1.009 & -0.117 \\ 
  % 43 & -0.095 & 0.116 & 0.645 & 0.980 & 1.009 & -0.055 \\ 
  % 44 & -0.101 & 0.125 & 0.692 & 0.980 & 1.009 & -0.068 \\ 
  % 45 & -0.085 & 0.117 & 0.651 & 0.980 & 1.008 & -0.086 \\ 
  % 46 & -0.092 & 0.117 & 0.649 & 0.970 & 1.008 & -0.048 \\ 
  % 47 & -0.095 & 0.115 & 0.638 & 0.980 & 1.007 & -0.069 \\ 
  % 48 & -0.096 & 0.112 & 0.620 & 0.970 & 1.009 & -0.076 \\ 
  % 49 & -0.090 & 0.108 & 0.601 & 0.980 & 1.009 & -0.083 \\ 
  % 50 & -0.087 & 0.119 & 0.662 & 0.990 & 1.009 & -0.029 \\ 
   \hline
\end{tabular}
\caption{Simulation and convergence diagnostics over 1000 MCMC simulations.}
\label{subsetresults}
\end{table}

From the simulation diagnostics it appears that as little as three iterations is sufficient to draw valid inference. Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate 20-30 iterations. Apparently, they are not perfect for MI data [move interpretation to Discussion!].

Add simulation time (computational cost) to table? Add empirical SE of all diagnostics? Sample some random simulations and check distribution and plausibility of imputations?

@@

This note illustrates that conventional convergence diagnostics behave differently on MI data as compared to the output of 'regular' MCMC algorithms. The recommended threshold for $\widehat{R}$ is too stringent for MI data.

Moreover, $\widehat{R}$ could theoretically not be smaller than one, yet it happened several times in this study (see online appendix XYZ). How could $\widehat{R}$ smaller than 1 occur? The number of simulations is smaller than in `regular' MCMC processes. Therefore, the '$(n-1/n)$' correction factor can influence the estimated potential scale reduction factor. This downwards bias is in the opposite direction than expected: ``The mixture-of-sequences variance, V, should stabilize as a function of n. (Before convergence, we expect $\sigma^2$ to decrease with n, only increasing if the sequences explore a new area of parameter space, which would imply that the original sequences were not overdispersed for the particular scalar summary being monitored.)'' \cite[p~438]{broo98}.
